On the Bit Complexity of Finding Points in Connected Components of a Smooth Real Hypersurface

### 2020-02-01 main.tex
Update main.tex
Updated Section 8: Proof of H'_i(7), quantitative Noether Normalization 
- I still do not understand (fully) how to apply Cayley H. However, I think I have everything else done in this section.

\usepackage{mathrsfs}
% new commands and shortcuts
\def\bs{\textit{\textbf{s}}}
\def\Lb{\textit{\textbf{L}}}
\def\Xb{\textit{\textbf{X}}}
\def\xb{\textit{\textbf{x}}}
\def\fb{\textit{\textbf{f}}}
\def\ab{\textit{\textbf{a}}}
\def\ajb{\overline{a_j}}
\def\bjb{\overline{b_{j,l}}}
\def\pjb{\overline{P_{j}}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\htt}{ht}
\def\C{\mathbb{C}}
\def\vt{\vartheta}
\def\Q{\mathbb{Q}}
\def\R{\mathbb{R}}
\def\rs{\mathscr{R}}
\def\P{\mathscr{P}}
\def\d{\delta}
\def\sing{ \textrm{sing}}
\def\codim{ \textrm{codim}}
\def\jac{ \textrm{jac}}
\def\exp{\textrm{exp}}
\def\grad{\textrm{grad}}
\def\rank{\textrm{rank}}
\def\reg{\textrm{reg}}
\def\jt{\widetilde{J}}
\def\At{\widetilde{A}}
\def\Yt{\widetilde{Y}}
\def\dt{\widetilde{d}}
\def\Dt{\widetilde{D}}
\def\rk{\textrm{rank }}
\def\pa{\partial}
\def\D{\Delta}
\def\Z{\frak{Z}}
\def\I{\frak{I}}
\def\Is{\frak{I}^{\star}}
\def\A{\frak{A}}
\def\fp{\frak{P}}
\def\sp{\mathscr{P}}
\def\la{\langle}
\def\ra{\rangle}
% matrices 
\def\bbm{\begin{bmatrix}}
\def\ebm{\end{bmatrix}}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
 We present a full analysis of an efficient algorithm for the
 computation of at least one point in each connected component of a
 smooth real hypersurface, taking into account bit complexity and
 all coefficient growth.  This problem is a basic and important
 subroutine used in semi-algebraic geometry. For instance, it is used
 in determining an upper bound on the number of connected components
 of a real hypersurface. The algorithm requires some genericity
 properties, which we ensure through random changes of variables that
 produce generic coordinate systems. In order to ensure these
 properties with arbitrarily high probability, we developed a new
 quantitative extension of Thom's weak transversality theorem, and a
 new quantitative Noether normalization statement for polar
 varieties. These statements enable us to give a precise probability
 analysis of a monte carlo algorithm, where we can guarantee
correctness with arbitrarily high probability.
\par 
  Algorithms for deciding connectivity queries in real algebraic
  geometry have been developed in a series of papers
  \cite{a,b,c,d}. However, these algorithms are based in the
  arithmetic complexity model, and analyses of the bit complexity
  questions were left for future work.  Our results are an important
  first step in this analysis.
\end{abstract}

@@

Let $\A$ be an $n \times n$ matrix of new indeterminates $(\A_{i,j})_{1 \leq i , j \leq n}.$ Define $f^{\A} \in \mathbb{Q}(\A)[\textbf{X}]$ as $f(\A\textbf{X})$ and $V^\A=V(f^\A).$ Recall from Section 4.1.4 that $f^\A$ satisfies $\textbf{H}_i^{'}(1),\textbf{H}_i^{'}(2),$ and $\textbf{H}_i^{'}(3).$ Throughout this section, for simplicity let $\I^\A$ denote $\I(\pi_i,V^\A)$ and let $W^\A$ denote $W(\pi_i,V^\A).$ To simplify notation, we will let $i \in \{1,\hdots,\}$ be fixed, and we wil avoid writing $i$ where the context is clear. 

@@

Let $f \in \mathbb{Q}[X_1,\hdots,X_n]$ be squarefree with total degree $D$ and $V(f)$ smooth. Recall $\textbf{H}_i^{'}(4):$ \textit{$W(\pi_i,V)$ is in Noether position for $\pi_{i-1}$}, which is equivalent to having the extension  

@@

@@

Section 8 is devoted to proving Theorem 8.1. The majority of work is in upper bounding the degree of $\Delta_3,$ as in \cite{EMP} it was shown that $f^A$ satisfies $\textbf{H}_i^{'}$ for a generic change of variables $A\in\C^{n^2}.$ 

@@

then $A$ cancels none of the denominators of $a_{j}$ or $b_{j,l}$. Define $h_j \in \Q[\A]$ to be the product of the denominators of the coefficients of $a_j$ and $b_{j,l}$. Hence, if $A \not \in V(h_j)$ then $A$ cancels none of the denominators $a_{j}$ or $b_{j,l}$ which are providing the integral dependence relation. Define

@@

\Xi_i := V(h_i,\hdots,h_n).

@@

For each $j \in \{i,\hdots,n\}$, we want a bound on the degree of $h_j$. Then by applying Bezout's theorem we can obtain a bound on the degrees of $\Xi_i$. We can obtain bounds for $h_j$ by applying an effective Nullstellensatz theorem. However, the theorem requires $P_{j}$ to be a polynomial. We can obtain a polynomial from $P_{j}$ by clearing denominators. We can multiply $P_{j}$ by $d_j \in \Q[\A_{i,j}]$ equal to the least common multiple of the denominators of $P_{j}.$ Put 

@@

$\deg \pjb\leq (D+1)^{n-i+1}.$ (The same for all $j$?)

@@

Let $\Is \subset \Q[\A,\Xb]$ be the extension of $\I$ in  $\Q[\A,\Xb],$ and consider the finite field extension

@@

As $\pjb$ is monic (it is not....), the minimal polynomial of the linear mapping defined by multiplication by $X_j$ in $\Q(\A,\Xb_{i-1})$ is equal to $\pjb$. It therefore follows by \cite[Proposition 1]{CGR} that

@@

\deg d_j \leq \deg \pjb \leq (D+1)^{n-i+1}.

@@

We have

@@

\deg_{\Xb}(1-T\pjb) \leq \deg \pjb +1 \leq (D+1)^{n-i+1}+1

@@

\deg_{\A}(1-T\pjb) \leq \deg \pjb\leq (D+1)^{n-i+1}.

@@

\alpha_j = \sum_{l=1}^{n-i+1} A_{j,l}(\textit{\textbf{X}},T)g_l + B_j(\textit{\textbf{X}},T)(1-\pjb T), \\ g_l \in 

@@

\right\},~ A_{j,l},B_j \in \mathbb{Q}[\A][\Xb][T]

@@

with the degree of $\alpha_j$ upper bounded by

@@

    &\left(\prod_{l=1}^{n-i+1} \deg_{\Xb,T} g_l\right)\deg_{\A} (1-T\pjb)\\ 
    + ~~~~~&(n-i+1)\left(\deg_{\Xb,T}\pjb\right)\left(\prod_{l=1}^{n-i+1}\deg_{\Xb,T} g_l \right)\deg_{\A} g_l\\
    \leq ~~~~~&D^{n-i+1}\left( (D+1)^{n-i+1}\right)\\
    +~~~~~&(n-i+1)\left( (D+1)^{n-i+1}+1\right)D^{n-i+2}.

@@

Now substitute $T \leftarrow 1/\pjb$ so that 

@@

&= \sum_{i=1}^{n_l} A(\A,\Xb,1/\pjb)\\
&= \sum_{i=1}^{n_l} \left(\frac{1}{P_j^{k_l}}\right)\At_{j,l,i}(\A,\textit{\textbf{X}}), ~k_l \in \mathbb{Z}.

@@

Now let $K := \max\{k_l\}$ so that 

@@

    &= \sum_{l=1}^{n-i+1}\left( \sum_{i=1}^{n_l}\left(\pjb^{K-k_{n_l}}\right)\At_{j,l,i}(\A,\Xb) \right)g_l \in \Q[\A][\Xb],

@@

which shows that $\alpha_j$ is the only remaining denominator (i.e.\ the only denominators are $d_j$ and $\alpha_j$). We know that $\alpha_j$ is not zero, we know the leading coefficient of $\overline{P_j}$ is $d_j$ which is not zero, we have an upper bound for the degrees of $\alpha_j$, and we know $\deg d \leq \deg P_j = (D+1)^{n-i+1}$, where $\pjb=d_jP_j.$ By making $\pjb$ monic (this is just $P_j$?), the integrality of 

@@

\mathbb{Q}(\frak{A}_{i,j})[\textbf{X}_{\leq i-1}]\rightarrow\mathbb{Q}(\frak{A}_{i,j})[\textbf{X}]/\I^{\A}

@@

holds with $\pjb(X_j)\in \I^{\A}.$

@@

\subsection{Proof of Theorem 5.1?}

@@

There is a Zariski-open subset $\Gamma \subset \GL(n,\C)$ such that for $A \in \Gamma$, the entries of $A$ cancel none of the denominators of the coefficients of the generators of $\I(\pi_i,V^{A}).$ Thus, we set

@@

\Xi := \C^{n^2} - \Gamma = V(h),

@@

with $h \in \Q[\A_{i,j}]$ defined as the least common multiple of the denominators of the generators of $\I^\A.$ Since $\pjb^k \in \I^\A$ and since
\pjb^k = \frac{1}{\alpha}\left( Q(\A_{i,j},\Xb) \right), Q \in \Q[\A_{i,j}][\Xb],
it follows that 
\deg \Xi = \deg h \leq \deg \alpha \leq D^{\frac{3}{2} n \ln n+1}.
Furthermore, since $\pjb^k \in \I^\A,$ it follows from by proposition 5.2 that, if $A \in \C^{n^2} - \Xi$ then the extension  
\C[\textbf{X}_{\leq i-1}]\rightarrow\C[\textbf{X}]/\frak{I}^\A
is integral, and thus $V(\I^A)=W^\A$ is in Noether position for $\pi_{i-1}.$ Therefore, for each $i \in\{1,\hdots,n\}, f^A$ satisfies $\textbf{H}_i^{'}(4)$.

>>>>

% test 
\usepackage{mathrsfs}
% new commands and shortcuts
\def\bs{\textit{\textbf{s}}}
\def\Lb{\textit{\textbf{L}}}
\def\Xb{\textit{\textbf{X}}}
\def\xb{\textit{\textbf{x}}}
\def\fb{\textit{\textbf{f}}}
\def\ab{\textit{\textbf{a}}}
\def\ajb{\overline{a_j}}
\def\bjb{\overline{b_{j,l}}}
\def\pjb{\overline{P_{j}}}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\htt}{ht}
\def\C{\mathbb{C}}
\def\vt{\vartheta}
\def\Q{\mathbb{Q}}
\def\R{\mathbb{R}}
\def\rs{\mathscr{R}}
\def\P{\mathscr{P}}
\def\d{\delta}
\def\sing{ \textrm{sing}}
\def\codim{ \textrm{codim}}
\def\jac{ \textrm{jac}}
\def\exp{\textrm{exp}}
\def\grad{\textrm{grad}}
\def\rank{\textrm{rank}}
\def\reg{\textrm{reg}}
\def\jt{\widetilde{J}}
\def\At{\widetilde{A}}
\def\Yt{\widetilde{Y}}
\def\dt{\widetilde{d}}
\def\Dt{\widetilde{D}}
\def\rk{\textrm{rank }}
\def\pa{\partial}
\def\D{\Delta}
\def\Z{\frak{Z}}
\def\I{\frak{I}}
\def\Is{\frak{I}^{\star}}
\def\A{\frak{A}}
\def\fp{\frak{P}}
\def\sp{\mathscr{P}}
\def\la{\langle}
\def\ra{\rangle}
% matrices 
\def\bbm{\begin{bmatrix}}
\def\ebm{\end{bmatrix}}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We present a full analysis of an efficient algorithm for the
  computation of at least one point in each connected component of a
  smooth real hypersurface, taking into account bit complexity and
  all coefficient growth.  This problem is a basic and important
  subroutine used in semi-algebraic geometry. For instance, it is used
  in determining an upper bound on the number of connected components
  of a real hypersurface. The algorithm requires some genericity
  properties, which we ensure through random changes of variables that
  produce generic coordinate systems. In order to ensure these
  properties with arbitrarily high probability, we developed a new
  quantitative extension of Thom's weak transversality theorem, and a
  new quantitative Noether normalization statement for polar
  varieties. These statements enable us to give a precise probability
  analysis of a monte carlo algorithm, where we can guarantee
  correctness with arbitrarily high probability.

  Algorithms for deciding connectivity queries in real algebraic
  geometry have been developed in a series of papers
  \cite{a,b,c,d}. However, these algorithms are based in the
  arithmetic complexity model, and analyses of the bit complexity
  questions were left for future work.  Our results are an important
  first step in this analysis.

\end{abstract}

@@

Let $\A$ be an $n \times n$ matrix of new indeterminates $(\A_{i,j})_{1 \leq i , j \leq n}.$ Define $f^{\A} \in \mathbb{Q}(\A)[\textbf{X}]$ as $f(\A\textbf{X})$ and $V^\A=V(f^\A).$ Recall from Section 4.1.4 that $f^\A$ satisfies $\textbf{H}_i^{'}(1),\textbf{H}_i^{'}(2),$ and $\textbf{H}_i^{'}(3).$ Throughout this section, to simplify notation, let $\I^\A$ denote $\I(\pi_i,V^\A)$ and let $W^\A$ denote $W(\pi_i,V^\A).$ Furthermore, throughout this section, consider that $i \in \{1,\hdots,n\}$ is fixed. We will avoid writing $i$ where the context is clear. 

@@

Let $f \in \mathbb{Q}[X_1,\hdots,X_n]$ have total degree $D$. Assume $f$ satisfies \textbf{H}, so that $f$ is squarefree with $V(f)$ smooth. Recall $\textbf{H}_i^{'}(7):$ \textit{$W(\pi_i,V)$ is in Noether position for $\pi_{i-1}$}, which is equivalent to having the extension  

@@

\C[\textbf{X}_{\leq i-1}]\rightarrow\C[\textbf{X}]/\frak{I}(\pi_i,V)

@@

Section 8 is devoted to proving Theorem 8.1. It has already been shown in \cite{EMP} that $f^A$ satisfies $\textbf{H}_i^{'}$ for a generic change of variables. The new contribution is in upper bounding the degree of $\Delta_3.$

@@

then $A$ cancels none of the denominators of $a_{j}$ or $b_{j,l}$. Define $h_j \in \Q[\A]$ to be the product of the denominators of the coefficients of $a_j$ and $b_{j,l}$. Hence, if $A \in \Q^{n^2}-V(h_j)$ then $A$ cancels none of the denominators of $a_{j}$ or $b_{j,l},$ which are providing the integral dependence relation. Now we define 

@@

\Xi_i := V(h_i,\hdots,h_n), 

@@

and we obtain a bound on the degree of $\Xi_i$ by applying Bezout's theorem. Hence, if $A\in \Q^{n^2}-\Xi_i$ then
P_j(X_j) \in \I^A,  
for each $j \in \{i,\hdots,n\},$ and therefore the extension 
\mathbb{C}[{\Xb}_{\leq i-1}]\rightarrow\mathbb{C}[\Xb]/\I(\pi_i,V^{A})
is integral. We can obtain bounds for $h_j$ by applying an effective Nullstellensatz theorem. However, the theorem requires $P_{j}$ to be a polynomial. We can obtain a polynomial from $P_{j}$ by clearing denominators. We can multiply $P_{j}$ by the least common multiple of its denominators, which we will denote by $d_j \in \Q[\A_{i,j}]$. Put 

@@

$\deg \pjb\leq (D+1)^{n-i+1}.$ 

@@

\textbf{Eric and Mark}: I am unsure of how to do this part correctly, so I will assume 
\deg \pjb\leq \Dt_j.
Let $\Is \subset \Q[\A,\Xb]$ be the extension of $\I^{\A}$ in  $\Q[\A,\Xb],$ and consider the finite field extension

@@

The minimal polynomial of the linear mapping defined by multiplication by $X_j$ in $\Q(\A,\Xb_{i-1})$ is \textbf{almost}(?) equal to $\pjb$ (the leading coefficient of $\pjb$ is $d_j$). It therefore follows by \cite[Proposition 1]{CGR} that

@@

\deg d_j \leq \deg \pjb \leq \Dt_j.

@@

To apply the effective Nullstellensatz theorem, we first need to determine degree bounds, separately in $\Xb$ and $\A,$ of the polynomials in the membership relationship. We have

@@

\deg_{\Xb,T}(1-T\pjb) \leq \Dt_j +1 

@@

\deg_{\A}(1-T\pjb) \leq \Dt_j.

@@

\alpha_j = \sum_{l=1}^{n-i+1} A_{j,l}(\A,\textit{\textbf{X}},T)g_l + B_j(\A,\textit{\textbf{X}},T)(1-\pjb T), \\ g_l \in 

@@

\right\},~ A_{j,l},B_j \in \mathbb{Q}[\A][\Xb][T],

@@

and with the degree of $\alpha_j$ upper bounded by

@@

    &\left(\prod_{k=1}^{n-i+1} D\right)\Dt_j 
    + (n-i+1)\left(\Dt_j+1\right)\left(\prod_{k=1}^{n-i}D \right)D\\
    = &\left(D^{n-i+1}\right)\Dt_j+
    (n-i+1)\left(\Dt_j+1\right)\left(D^{n-i+1} \right)\\
    &\leq D^{n}\Dt_j + n\left(\Dt_j+1\right)D^{n}\\
    &=D^n\left(\Dt_j + n\left(\Dt_j+1\right) \right).

@@

Now we will show that $\alpha_j$ is the only remaining denominator whose degree we need to bound. Substitute $T \leftarrow 1/\pjb$ so that 

@@

&= \sum_{i=1}^{n_l} A_{j,l,i}(\A,\Xb,1/\pjb)\\
&= \sum_{i=1}^{n_l} \left(\frac{1}{P_j^{k_i}}\right)\At_{j,l,i}(\A,\textit{\textbf{X}}), ~k_i \in \mathbb{Z}.

@@

Let $K := \max\{k_l\}$ so that 

@@

    &= \sum_{l=1}^{n-i+1}\left( \sum_{i=1}^{n_l}\left(\pjb^{K-k_{i}}\right)\At_{j,l,i}(\A,\Xb) \right)g_l \in \Q[\A][\Xb].

@@

which shows that $\alpha_j$ is the only remaining denominator whose degree we need to bound other than $d_j$. Note as well that
\pjb^K \in \I^{\A}.

@@

\deg h_j &\leq \left(\deg \alpha_j\right)\left(\deg d_j\right)\\
    &\leq \left(D^n\left(\Dt_j + n\left(\Dt_j+1\right) \right)\right)\Dt_j\\
    &=D^n\Dt_j\left(\Dt_j + n\left(\Dt_j+1\right) \right).

@@

\subsection{Proof of Theorem 8.1}

@@

We can apply B\'ezout and obtain

@@

    \deg \Xi_i = \deg V(h_i,\hdots,h_n)
    &\leq \prod_{j=i}^n \deg h_j\\
    &\leq  \prod_{j=i}^n \left( D^n\Dt_j\left(\Dt_j + n\left(\Dt_j+1\right) \right) \right) \\
    &=  \left( D^n\Dt_j\left(\Dt_j + n\left(\Dt_j+1\right) \right) \right)^{n-i+1}\\
    &\leq  \left( D^n\Dt_j\left(\Dt_j + n\left(\Dt_j+1\right) \right) \right)^{n}.
And therefore, by another application of B\'ezout, 
    \deg& \Delta_3 \leq \prod_{i=1}^n \deg \Xi_i\\
    &\leq \left( D^n\Dt_j\left(\Dt_j + n\left(\Dt_j+1\right) \right) \right)^{n^2},
    &(=D^{O(n^3 \log n)}\textrm{?})
where if $A \in \C^{n^2} - \Delta_3$ then, for each $i \in \{1,\hdots,n\},$ the extension 

@@

@@
